$p$-adic precision

### 2014-02-05 padicprec.tex
Reference for Somos

However, on the other hand, one can prove that the SOMOS 4 sequence 
exhibits the \emph{Laurent phenomenon}. This means that, for all integer 
$n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
From the latter formula, it follows directly that if $u_0$, $u_1$,
$u_2$ and $u_3$ are known up to precision $O(p^N)$ then all $u_n$'s
are also known with the same precision. Thus, the term $v_n$ that 
appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
precision but some numerical instability related to the algorithm ---
which includes the model precision --- we used to compute to SOMOS 
sequence.

@@

\begin{deftn}
Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
\emph{differentiable} at $v_0 \in U$ if there exists a 
continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
\subset U$ containing $v_0$ with:
\[ 
\forall v, w \in U_\varepsilon, \quad
\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
\leq \varepsilon \Vert v-w \Vert. 
\]
The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
\end{deftn}

@@

\begin{proof}
We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, by 
the open mapping Theorem, there exists $C>0$ such that $B_F(1) \subset 
C \cdot f'(0) \cdot B_E(1)$. Let $\varepsilon>0$ be such that 
$\varepsilon C < 1$. Let $U_\varepsilon \subset E$ be given by the fact 
that $f$ is differentiable at $0$. We may assume $U_\varepsilon = 
B_E(\delta)$ for some $\delta >0$.

Let $r \in (0, \delta)$.
Let us show that $f$ maps surjectively $H$ onto $f'(0) \cdot H$.
Let $x \in H$. We must show that $f(x) \in f'(0) \cdot H$. 
By differentiability at $0$, $\Vert f(x)-f'(0) \cdot x \Vert \leq 
\varepsilon \Vert x \Vert $. Therefore, there exists $y \in f'(0) \cdot 
H$ such that 
$\Vert y \Vert \leq \varepsilon r$. By the open mapping Theorem and the 
surjectivityy of $f'(0)$, there exists $x_1 \in E$ such that $f'(0) \cdot 
x_1 =y$ and $\Vert x_1 \Vert \leq \varepsilon C \cdot r < r$. Thus, $x_1 
\in B^-_E(r) \subset H$ and $f(x)= f'(0) \cdot (x-x_1) \in f'(0) 
\cdot H$.

@@

If we denote by $NP_F$ the Newton polygon of the power series 
\[
F(t) = \sum_{k\geq 0}^{+\infty} \log_p (\interleave f^{(k+1)}(0) \interleave ) t^k
\]
then it is enough that:
\[
\sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \left( \log_p l - \frac{1}{p-1} \right) \right) \geq - \log_p \varepsilon.
\]

@@

Another usual operations on polynomials is the factorization. Suppose 
that we are given a monic polynomial $P \in K[X]$ written as a product 
$P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. Hensel's 
lemma implies that there exists a small neighbourhood $\mathcal U$ of 
$P_0$ such that any monic polynomial $P \in \mathcal U$ factors uniquely 
as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ and $B_0$ 
respectively. Thus, we can consider the map $f : P \mapsto (A,B)$ 
defined on the subset of $\mathcal U$ consisting of monic polynomials.
We want to differentiate $f$ at $P_0$. For this, we differentiate the 
equality $P = A B$ around $P_0$, obtaining 
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

An important special case of the previous study occurs when $A_0$ has 
degree $1$, that is $A_0(X) = X - \alpha$ with some $\alpha \in K$. The 
map $P \mapsto A$ is then nothing but the mapping that follows 
the simple root $\alpha$. Of course, its differential around $P_0$ can 
be computed by the above method. Nevertheless there is a more direct way 
to obtain it in this particular case: we write $(P + dP)(\alpha + 
d\alpha) = 0$ and expand this. We obtain this way $P'(\alpha) d\alpha + 
dP(\alpha) = 0$ where $P'$ denotes the derivative of $P$. Since $\alpha$ 
is a simple root, $P'(\alpha)$ does not vanish and we recover $d \alpha 
= - \frac{dP(\alpha)}{P'(\alpha)}$.

@@

We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
variables over $K$ and fix a monomial order on it. We then have a 
notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ are polynomials
in $K[\XX]$, there exists a writing:
$$f = q_1 f_1 + \cdots + q_s f_s + r$$
where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
assumption, there still exists a canonical choice of $r$. We do not know
if the computation of the differential that follows extends to a more
general setting.} (of the ideal generated by these polynomials), the 
polynomial $r$ is uniquely determined and called the \emph{remainder} of 
the division of $f$ by the family $(f_1, \ldots, f_s)$. The map 
$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
the case of univariate polynomials, we can compute its differential: we 
find that $dr$ is obtained as the remainder of the division of $f - (q_1 
\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.

We now focus on the computation of Gr\"obner basis. Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
then consider the map sending a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{Vaccon}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
map at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA = 
(d g_1, \ldots, d g_t).$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d g_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.

@@

Let us now consider the open subset of $M_{n,n-d}(K)$ consisting of 
matrices of full rank. The left kernel defines a mapping $\text{LK} : 
M_{n,n-d}(K) \to \Grass(d,n)$. Let us prove that it is differentiable 
and compute its differential around some point $M \in M_{n,n-d}(K)$. Of 
course, there exists a neighbourhood $\mathcal U$ of $M$ whose image is 
entirely contained in a given chart $U_P$. We assume for simplicity that 
$P$ is the identity. On $\mathcal U$, the map $\text{LK}$ corresponds in 
our chart to the map that takes $M$ to the unique matrix $N \in
M_{d, n-d}(K)$ such that $\begin{pmatrix} I_d & N \end{pmatrix} \cdot M 
= 0$. The implicit function theorem then implies that $\text{LK}$ is
differentiable. Furthermore, its differential satisfies the relation
$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
from which we can compute $dN$ by projecting on the $(n-d)$ last
columns.

@@

Recall (\emph{cf} \S \ref{ssec:stepbystep}) that a SOMOS 4 sequence is a 
four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 
+ u_{n+3}^3}{u_n}$. We recall also that SOMOS sequences exhibit the 
Laurent phenomenon: its means that the four initial terms $u_0, u_1, 
u_2, u_3$ are some indeterminates, that each $u_n$ is a Laurent 
polynomial with coefficients in $\Z$ in these indeterminates.

@@

>>>>

However, on the other hand, one can prove that the SOMOS 4 sequence 
exhibits the \emph{Laurent phenomenon}. This means that, for all integer 
$n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$ (see \cite{Fomin-Zelevinsky}).
From the latter formula, it follows directly that if $u_0$, $u_1$,
$u_2$ and $u_3$ are known up to precision $O(p^N)$ then all $u_n$'s
are also known with the same precision. Thus, the term $v_n$ that 
appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
precision but some numerical instability related to the algorithm ---
which includes the model precision --- we used to compute to SOMOS 
sequence.

@@

\begin{deftn} \label{deftn : diff}
Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
\emph{differentiable} at $v_0 \in U$ if there exists a 
continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
\subset U$ containing $v_0$ with:
\[ 
\forall v, w \in U_\varepsilon, \quad
\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
\leq \varepsilon \Vert v-w \Vert. 
\]
The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
\end{deftn}

@@

\begin{proof}
We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, by 
the open mapping Theorem, there exists $C>0$ such that $B_F(1) \subset 
C \cdot f'(0) \cdot B_E(1)$. Let $\varepsilon>0$ be such that 
$\varepsilon C < 1$. Let $U_\varepsilon \subset E$ be given by definition \ref{deftn : diff} and the fact 
that $f$ is differentiable at $0$. We may assume $U_\varepsilon = 
B_E(\delta)$ for some $\delta >0$.

Let $r \in (0, \delta)$.
Let us show that $f$ maps surjectively $H$ onto $f'(0) \cdot H$.
Let $x \in H$. We must show that $f(x) \in f'(0) \cdot H$. 
By differentiability at $0$, $\Vert f(x)-f'(0) \cdot x \Vert \leq 
\varepsilon \Vert x \Vert $. Therefore, there exists $y \in f'(0) \cdot 
H$ such that 
$\Vert y \Vert \leq \varepsilon r$. By the open mapping Theorem and the 
surjectivity of $f'(0)$, there exists $x_1 \in E$ such that $f'(0) \cdot 
x_1 =y$ and $\Vert x_1 \Vert \leq \varepsilon C \cdot r < r$. Thus, $x_1 
\in B^-_E(r) \subset H$ and $f(x)= f'(0) \cdot (x-x_1) \in f'(0) 
\cdot H$.

@@

If we denote by $NP_F$ the Newton polygon application of the power series 
\[
F(t) = \sum_{k\geq 0}^{+\infty} \log_p (\interleave f^{(k+1)}(0) \interleave ) t^k
\]
then it is enough that:
\[
\sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \left( \log_p l - \frac{1}{p-1} \right) \right) \geq - \log_p \varepsilon.
\]

@@

Another usual operations on polynomials is the factorization. Suppose 
that we are given a monic polynomial $P \in K[X]$ written as a product 
$P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. Hensel's 
lemma implies that there exists a small neighbourhood $\mathcal U$ of 
$P_0$ such that any monic polynomial $P \in \mathcal U$ factors uniquely 
as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ and $B_0$ 
respectively. Thus, we can consider the application $f : P \mapsto (A,B)$ 
defined on the subset of $\mathcal U$ consisting of monic polynomials.
We want to differentiate $f$ at $P_0$. For this, we differentiate the 
equality $P = A B$ around $P_0$, obtaining 
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

An important special case of the previous study occurs when $A_0$ has 
degree $1$, that is $A_0(X) = X - \alpha$ with some $\alpha \in K$. The 
application $P \mapsto A$ is then nothing but the mapping that follows 
the simple root $\alpha$. Of course, its differential around $P_0$ can 
be computed by the above method. Nevertheless there is a more direct way 
to obtain it in this particular case: we write $(P + dP)(\alpha + 
d\alpha) = 0$ and expand this. We obtain this way $P'(\alpha) d\alpha + 
dP(\alpha) = 0$ where $P'$ denotes the derivative of $P$. Since $\alpha$ 
is a simple root, $P'(\alpha)$ does not vanish and we recover $d \alpha 
= - \frac{dP(\alpha)}{P'(\alpha)}$.

@@

We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
variables over $K$ and fix a monomial order on it. We then have a 
notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ are polynomials
in $K[\XX]$, there exists a writing:
$$f = q_1 f_1 + \cdots + q_s f_s + r$$
where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
assumption, there still exists a canonical choice of $r$. We do not know
if the computation of the differential that follows extends to a more
general setting.} (of the ideal generated by these polynomials), the 
polynomial $r$ is uniquely determined and called the \emph{remainder} of 
the division of $f$ by the family $(f_1, \ldots, f_s)$. The application 
$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
the case of univariate polynomials, we can compute its differential: we 
find that $dr$ is obtained as the remainder of the division of $f - (q_1 
\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.

We now focus on the computation of Gr\"obner basis. Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{Vaccon}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
application at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA = 
(d g_1, \ldots, d g_t).$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d g_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.

@@

Let us now consider the open subset of $M_{n,n-d}(K)$ consisting of 
matrices of full rank. The left kernel defines a mapping $\text{LK} : 
M_{n,n-d}(K) \to \Grass(d,n)$. Let us prove that it is differentiable 
and compute its differential around some point $M \in M_{n,n-d}(K)$. Of 
course, there exists a neighbourhood $\mathcal U$ of $M$ whose image is 
entirely contained in a given chart $U_P$. We assume for simplicity that 
$P$ is the identity. On $\mathcal U$, the map $\text{LK}$ corresponds in 
our chart to the application that takes $M$ to the unique matrix $N \in
M_{d, n-d}(K)$ such that $\begin{pmatrix} I_d & N \end{pmatrix} \cdot M 
= 0$. The implicit function theorem then implies that $\text{LK}$ is
differentiable. Furthermore, its differential satisfies the relation
$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
from which we can compute $dN$ by projecting on the $(n-d)$ last
columns.

@@

Recall (\emph{cf} \S \ref{ssec:stepbystep}) that a SOMOS 4 sequence is a 
four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 
+ u_{n+3}^3}{u_n}$. We recall also that SOMOS sequences exhibit the 
Laurent phenomenon: its means that the four initial terms $u_0, u_1, 
u_2, u_3$ are some indeterminates, that each $u_n$ is a Laurent 
polynomial with coefficients in $\Z$ in these indeterminates (see \cite{Fomin-Zelevinsky}).

@@

\bibitem{Fomin-Zelevinsky}
  S.~Fomin, A.~Zelevinsky
  \emph{The Laurent phenomenon,}   Advances in Applied Mathematics 28: 119–144 (2002).
